\chapter*{Acknowledgements}
\markboth{Acknowledgements}{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}
% put your text here
Words are not merely enough to express my deep appreciation towards all the family members, friends, and mentors who unconditionally supported me throughout the past years. I feel exceedingly lucky to be surrounded with so many loving, caring, kind, and generous people. To be this lucky is exceptionally rare. All I can do to repay what I have received is to try to make someone feel as lucky as I felt.


First and foremost, I would like to express my sincere gratitude toward my advisor prof. Jan Hesthaven for his excellent mentorship. I would like to thank him for his continuous support and encouragement, as well as his complete faith in me. As you mentioned once, scientific research is only a fraction of what is to learn in this journey. Indeed, what you have taught me is a worldview; a treasure that I will hold on to for the rest of my life.

I would like to thank prof. Bernard Haasdonk, for hosting me at Stuttgart University. Our brainstorming meeting shaped some of the main parts of this thesis. My sincere gratitude to my collaborators Dr. Ashish Bhatt, Nicol\'o Ripamonti, and Dr. Qian Wang. This thesis would not have been possible without them. Furthermore, I would like to thank SNSF and prof. Karen Willcox for providing me with the unique opportunity to be an exchange graduate student at MIT.



I gratefully acknowledge prof. Marc Troyanov, prof. Fabio Nobile, prof. Mario Ohlberger, and Dr. Kevin Carlberg for serving as my committee members. I would like to express my special appreciation for Dr. Kevin Carlberg's effort in reading my thesis in remarkable detail. His comments and suggestions significantly enhanced the quality and the credibility of my thesis. I also would like to extend my gratitude towards prof. Troyanov for giving a crash course on differential geometry. This, as well, largely contributed to the quality of this thesis. 


This work would have been truly impossible without the constant moral and emotional support of Claudia. She was always there for me, patiently and unconditionally, to help me get through difficulties in life, as well as in my research in mathematics. She always put me before her, and for this, I will forever be in dept.


I greatly appreciate the support received from my uncle Amin and my aunt Dorothe. They taught me to be dedicated, hard-working, and determined toward my career, but also to be a kind, generous, and a giving human being. They always welcomed me with open arms and made me feel at home with the delicious Persian food. They set an excellent example for me in life.


I am grateful to my parents, Ali and Homa, who never stopped providing me with love and moral support. Without this support, I wouldn't have gone so far. I would also like to express my gratitude to my brother, Heydar, and his family, Greta, Adrian, and Elmira, for never failing to cheer me up. I would also thank all the other family members who never stopped believing in me and supporting me throughout my life.


I am also grateful to all my friends in the USA. To Fran\c{c}ois for his infinite generosity and warm heart, to Ruzbeh for the amazing and memorable road trip in California, to Cory and Judy for teaching me how to fight my fears and for every enjoyable second of climbing and skiing with them, to Iman for hosting me in Austin and always being my best friend.


My special gratitude toward Paolo and Sebastian, for all the hikes, climbs, and trips we had together. I always looked upon them as older brothers.


I also thank all my amazing friends and colleagues in Switzerland.  I would like to extend my sincere gratitude to Delphine Vieira for always finding the best and the fastest solution for any issue I was facing at EPFL. Special thanks to Sylvie, for teaching me French and being there for me if I needed consolation. Many thanks to Allan, for going through all the ups and downs with me and showing me the way from one step ahead. Many thanks to Caterina, Fabian, Giacomo, Niccol\'o, Luca, Edoardo, Ana, Eva and all the other members of MCSS group and Mathicse group for making every second of my time at EPFL stimulating, exciting, and enjoyable. Huge thanks to my amazing Iranian friends Hossein and Parisa for being there for me from day one. Thanks to Marieke, Love, Franziska, Alexandra, and Konstantinos for all the memorable ski trips. I also thank Baptist, Fabrizio, Theodor, and Jon for being amazing mountain companions and trusting me with their lives. Last but not least, I would like to thank all of the climbing community in Lausanne and in Boston for always welcoming me with open arms and with positive attitude whether I was happy or sad, fresh or tired, fit or weak, or healthy or injured.

\bigskip
 
\noindent\textit{Lausanne, Oct 1st 2018}
\hfill B.~Maboudi
